---
title: "Reproducible Research: Peer Assessment 2"
author: "Dipanshu Verma"
date: "December 03, 2020"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = FALSE)
```

# Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities
and municipalities. Many severe events can result in fatalities, injuries, and property damage, and
preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm
database. This database tracks characteristics of major storms and weather events in the United States,
including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

# Overview
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about
severe weather events.

1. Across the United States, which types of events (as indicated in the **EVTYPE** variable)
are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

# Import Libraries

```{r}
library(ggplot2)
library(plyr)
library(gridExtra)
library(grid)
```

# Processing the data

```{r}
if(!file.exists("./storm_dataset")){
  dataset<-download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2","./storm_dataset")
}

storm  <- read.csv(bzfile("storm_dataset"))
```


```{r}
# number of unique event types
length(unique(storm$EVTYPE))
# translate all letters to lowercase
event_types <- tolower(storm$EVTYPE)
# replace all punct. characters with a space
event_types <- gsub("[[:blank:][:punct:]+]", " ", event_types)
length(unique(event_types))
# update the data frame
storm$EVTYPE <- event_types
```

# Find events that caused most death and injury

```{r}
casualties <- ddply(storm, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))
fatal_events <- head(casualties[order(casualties$fatalities, decreasing = T), ], 10)
injury_events <- head(casualties[order(casualties$injuries, decreasing = T), ], 10)
```

# Find events that caused most economic damage, but first adjust the variables according to the orders of magnitude described in the variables 
```{r}
exp_transform <- function(e) {
    # h -> hundred, k -> thousand, m -> million, b -> billion
    if (e %in% c('h', 'H'))
        return(2)
    else if (e %in% c('k', 'K'))
        return(3)
    else if (e %in% c('m', 'M'))
        return(6)
    else if (e %in% c('b', 'B'))
        return(9)
    else if (!is.na(as.numeric(e))) # if a digit
        return(as.numeric(e))
    else if (e %in% c('', '-', '?', '+'))
        return(0)
    else {
        stop("Invalid exponent value.")
    }
}
```

```{r cache=TRUE}
prop_dmg_exp <- sapply(storm$PROPDMGEXP, FUN=exp_transform)
storm$prop_dmg <- storm$PROPDMG * (10 ** prop_dmg_exp)
crop_dmg_exp <- sapply(storm$CROPDMGEXP, FUN=exp_transform)
storm$crop_dmg <- storm$CROPDMG * (10 ** crop_dmg_exp)
```


```{r}
# Compute the economic loss by event type
econ_loss <- ddply(storm, .(EVTYPE), summarize,
                   prop_dmg = sum(prop_dmg),
                   crop_dmg = sum(crop_dmg))
# filter out events that caused no economic loss
econ_loss <- econ_loss[(econ_loss$prop_dmg > 0 | econ_loss$crop_dmg > 0), ]
prop_dmg_events <- head(econ_loss[order(econ_loss$prop_dmg, decreasing = T), ], 10)
crop_dmg_events <- head(econ_loss[order(econ_loss$crop_dmg, decreasing = T), ], 10)
```

# Results and Plots

# Top dangerous weather event

```{r}
# Set the levels in order
p1 <- ggplot(data=fatal_events,
             aes(x=reorder(EVTYPE, fatalities), y=fatalities, fill=fatalities)) +
    geom_bar(stat="identity") +
    coord_flip() +
    ylab("Total number of Fatalities") +
    xlab("Event type") +
    theme(legend.position="none")
p2 <- ggplot(data=injury_events,
             aes(x=reorder(EVTYPE, injuries), y=injuries, fill=injuries)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    ylab("Total number of Injuries") +
    xlab("Event type") +
    theme(legend.position="none")
grid.arrange(p1, p2, top="Top deadly weather events")
```

# Economics Damage by most severe weather events
```{r}
# Set the levels in order
p1 <- ggplot(data=prop_dmg_events,
             aes(x=reorder(EVTYPE, prop_dmg), y=log10(prop_dmg), fill=prop_dmg )) +
    geom_bar(stat="identity") +
    coord_flip() +
    xlab("Event type") +
    ylab("Property damage in dollars (log-scale)") +
    theme(legend.position="none")
p2 <- ggplot(data=crop_dmg_events,
             aes(x=reorder(EVTYPE, crop_dmg), y=crop_dmg, fill=crop_dmg)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    xlab("Event type") +
    ylab("Crop damage in dollars") + 
    theme(legend.position="none")
grid.arrange(p1, p2, top="Weather costs to the US economy")
```



# Summary of Conclusions

- Tornado are responsible for the maximum number of fatalities and injuries.
- Droughts are responsible for maximum crop damage.
- Floods are responsible for maximum property damage.